\Title

### 2012-12-06 synmut.tex
major rearrangements, rewriting -> first draft from RAN, needs fleshing out, new figures etc

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt,a4paper,notitlepage,onecolumn]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{(Rise and fall of synonymous mutations)}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym}
\usepackage{color}
\usepackage{subfig}
\usepackage{graphicx}
%\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

@@

@@

HIV evolves rapidly within a single host during the course of the infection. The
driving forces shaping this process are the high mutation rate and the strong
selection imposed by the host immune system via killer T cells (CTLs) and
neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996}.

When the host develops a CTL or AB respose against a particular viral epitope,
rare HIV variants carrying mutated versions of the epitope, called {\it escape
mutants}, acquire a fitness advantage and spread rapidly in the viral
population, within a few months (see \figurename~\ref{fig:aft}, solid lines).
During chronic infection, the (Malthusian) effect size of this beneficial
mutations is of the order of $s_e \sim 0.01$~\citep{neher_recombination_2010}.
The viral \env{} gene shows the fastest rates of adaptation, because is both
rich of CTL epitopes and targeted by antibodies; its sequence diverges at rates
of the order of $1\%$ per year~\citep{shankarappa_consistent_1999}.

Many nucleotide polymorphisms that reach high frequency in the viral population
are escape mutations: all of them are {\it nonsynonymous} (i.e. they appear in
protein coding regions and change the amino acid sequence). Nonetheless,
nucleotide changes unrelated to immune escape are not rare, in \env{} and
elsewhere, and some of them become abundant alike, often rapidly. In particular,
it is not uncommon for {\it synonymous} mutations to reach frequencies of order
one within months from their first appearance (see \figurename~\ref{fig:aft},
dashed lines). The biological function of these mutations within the economy of
HIV is not well understood. By definition, the immunological phenotype, which is
decided at the protein level, is unaffected, but other biological and ecological
aspects of the viral lifestyle might be involved. In practice, a couple of
RNA-level phenotypes are known. For example, within \env{} a certain RNA
sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}.

Crucially for evolutionary purposes, the minor phenotypes caused by synonymous
mutations might have an effect on viral fitness. For instance, recent studies
have shown that genetically engineered HIV strains with skewed codon usage bias
(CUB) patterns towards more or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. In
this study, we try to characterize the fitness effects of synonymous
polymorphisms that, at some point during chronic infection, become abundant in
the viral population.

One simple way to assess the neutrality of synonymous mutations is to look at
their level of conservation. Deleterious mutations at functional sites are
expected to be absent or rare across the viral population; vice versa, mutant
alleles that reach high frequencies are expected to be neutral. If genetic sites
are independent, the equilibrium frequency of a deleterious allele with fitness
$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
neutral alleles have no equilibrium frequency and can slowly fix via genetic
drift~\citep{ewens_mathematical_2004}. If the focal synonymous mutant is linked
to another nonneutral allele, however, its frequency is the result of the
combined fitness effects of both sites, and simple conservation-level analyses
fail. Since recombination in HIV is known to happen
rarely~\citep{neher_recombination_2010, batorsky_estimate_2011}, the genetic
context of the synonymous change at hand must be taken into account. Our
results underline the importance of the latter scenario for intrapatient HIV
evolution.

\section{Results}
We analyze from time series of viral nucleotide sequences from single patients,
which span several years of chronic
infection~\citep{shankarappa_consistent_1999, bunnik_autologous_2008,
liu_selection_2006}. Plotting the allele frequencies against time for all
polymorphic sites, it is evident that, although nonsynonymous changes are
widespread, synonymous ones are also present in various occasions at high
frequency (see \figurename~\ref{fig:aft}).
\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Allele frequency trajectories of typical patient, C2-V5, nonsynonymous
(solid) and synonymous mutations (dashed lines). Most synonymous mutations are
not fixed. Colors are set according to the position of the site along the C2-V5
region (red to blue). Data from Ref.~\cite{shankarappa_consistent_1999}.}
\label{fig:aft}
\end{center}
\end{figure}
Moreover, many mutations appear to change rapidly in frequency as a flock,
especially at the lower end of the frequency spectrum. These observations lead
to the idea that linkage might be widespread at the time scale of the typical
selective sweep, i.e. a few months~\citep{neher_recombination_2010}. It seems
therefore likely that synonymous mutations are not attaining high frequencies by
genetic drift but are rather exponentially amplified by selection on linked
nonsynonymous sites, a process known as {\it genetic
draft}~\citep{neher_genetic_2011}.

To test this hypothesis, we focus on the long-time fate of such synonymous
derived alleles {\it after} they are already abundant. According to the neutral
theory of independent sites, (a) an allele starting at frequency $\nu$ will
reach fixation, in the long-time limit, with a probability $P_f(\nu) = \nu$, and
get extinct otherwise; (b) the time required to reach either boundary, fixation
or extinction, is essentialy the inverse of the population size (in
generations), hence much longer than five years for HIV~\citep{boltz_ultrasensitive_2012}.
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}}
\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions}}
\caption{Left panel: fixation probability of derived synonymous alleles is strongly
suppressed in C2-V5 versus other parts of the {\it env} gene, and of
nonsynonymous ones.
Right panel: especially hard is fixation of new alleles in conserved regions flanking the V
loops. The black dashed line is the prediction from neutral
theory, for comparison purposes. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

The left panel of \figurename~\ref{fig:fixp} shows the fixation probability of
derived alleles seen within various frequency windows. With regard to synonymous
alleles, the average over the whole HIV genome (organce line) matches closely
the expected result from neutral theory (dashed line). A different picture
emerges when the most variable part of the genome, V1-C5 region of \env{},
is zoomed into (red line). Two independent datasets, from
refs.~\citep{shankarappa_consistent_1999, bunnik_autologous_2008} yield the same
result, i.e. that synonymous mutations hardly fix at all.

A related observation can be made on the nonsynonymous alleles (blue line).
Although we see no difference between different parts of the HIV genome, the
neutral-like shape of $P_f$ is at odds with the na\"ive expectation that most
nonsynonymous alleles reaching high frequency are to-be-fixed escape mutations.

Moreover, as illustrated in \figurename~\ref{fig:fixtimes}, the time for
synonymous alleles to reach the first boundary starting from intermediate
frequencies is of the order of 500 days which, assuming a conservative
generation time of one day~\citep{perelson_hiv-1_1996}, correspond to the same
number of generations. This time would only be possible in a neutral scenario
if the population size of HIV were extremely small, of the order of $10^3$, much
smaller then both theory and experiments
indicate~\citep{boltz_ultrasensitive_2012}.
% WE SHOULD REALLY CITE OURSELVES IN THE (HOPEFUL) FUTURE HERE :-)
\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
\caption{Fixation or extinction times for synonymous alleles starting from
intermediate frequencies. The colored bands are the final fixation
probabilities expected from neutral theory; the observed alleles are fixed
less frequently than expected. The timescale of fixation/extinction is
approximately 500 days, corresponding to a selective effect of $\sim -0.001$.}
\label{fig:fixtimes}
\end{center}
\end{figure}

Taken together, these findings suggest that synonymous (and some nonsynonymous)
alleles are not propagating independently by genetic drift, but rather
hitchhiking on neighbouring beneficial alleles. This conclusion is also
consistent with the currently accepted estimate for the outcrossing rate of HIV,
of the order of $1\%$ per generation. Although synonymous alleles can be drafted
to high frequency by linkage and selection, they decouple on a time scale of the
order of 100 days, which is also the average time it takes a (nonsynonymous)
selected allele to spread through the population during chronic
infection~\citep{neher_recombination_2010}.

\figurename~\ref{fig:fixp} (left panel) also suggests that at least some
synonymous mutations in the V1-C5 region should be deleterious. We analyze this
hypothesis in two ways. First of all, we look for possible biological causes for
such a behaviour. Furthermore, we perform computer simulations of viral
evolution under a wide range of parameters in order to reproduce the results for
the fixation probability and study the regions of the parameter space compatible
with the observations.

One possible {\it a priori} explanation for deleterious fitness effects of
synonymous mutations is the presence of secondary structures in the viral RNA.
If any RNA secondary structures are functionally relevant for HIV replication,
mutations in nucleotides involved in those base pairs are expected to revert
more easily than by pure chance, to keep those structures in place. The
propensity of nucleotides to form base pairs in HIV has been measured
recently\citep{watts_architecture_2009}, and the HIV genome has resulted to be
highly structured~\citep{forsdyke_reciprocal_1995, watts_architecture_2009}.
Watts {\it et al.} have measured the propensity of any nucleoside to be attacked by
external chemicals (SHAPE reactivity) and shown that a higher reactivity
corresponds to a smaller likelihood of being involved in an RNA helix. As far as
the V1-C5 region is concerned, it has been shown that the short C1-5 regions
have a high tendency to form {\it insulating
stems}~\citep{watts_architecture_2009, sanjuan_interplay_2011}. Since the V1-5
loops have to be highly variable to ensure immune escape, their destabilizing
contribution to the viral RNA structure is though to be buffered by the
surrounding, base-pair rich stems. To test whether or not insulating stems might
be the reason for the reduced fixation probability, we partition the data
depending on their position in the V1-C5 region. Consistently with the
expectations, $P_f$ is more heavily readuced in the stems than in
the loops (see \figurename~\ref{fig:fixp}, right panel).

Despite this indirect result supporting the involvement of RNA secondary
structure in the deleterious effects of synonymous mutations in V1-C5, we set
out to seek more direct evidence. We partition all synonymous alleles observed
at intermediate frequencies above 10-15\% depending on their final destiny
(fixation or extinction). Subsequently, we align our sequences to the reference
NL4-3 strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
reactivities. As shown in \figurename~\ref{fig:SHAPE} (left panel) in a
cumulative histogram, the reactivity of fixed alleles are systematically larger
than of alleles that are doomed to extinction. In other words, alleles that are
likely to be breaking RNA helices are also more likely to revert and finally be
lost from the population.
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_nonVandflanking}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

Since the insulating stems are not the only RNA secondary structures found by
the authors of ref.~\citep{watts_architecture_2009}, we repeated the same
partitioning based on fate on the rest of the HIV genome. As shown in the right
panel of \figurename~\ref{fig:SHAPE}, there is no evidence that a similar
mechanism is acting outside of the V1-C5 region. This is not fully surprising,
because it is well known that single stranded RNA chains tend to form a lot of
spontaneous structures, most of which have no biological function. Furthermore,
the averaging over many different genomic regions might worsen the
signal-to-noise ratio, which is already quite low, and result in
indistinguishable histograms.

@@

We also perform computer simulations of evolving viral population under
selection and rare recombination. For this purpose, we use the recently
published package FFPopSim, which includes a module dedicated to intrapatient
HIV evolution~\citep{zanini_ffpopsim:_2012}. We analyze many combination of
parameters such as population size, recombination rate, selection coefficient
and density of escape mutations, epitope length, deleterious effect of
synonymous mutations. The purpose of the simulations is to probe the population
genetics mechanisms required to reproduce the fixation probabilities of
\figurename~\ref{fig:fixp}, simultaneously for synonymous and nonsynonymous
alleles.

The main result of the simulations is that genetic draft can indeed generate a
depression in fixation probabilities, but only under relatively tight
circumstances. In terms of model confidence, this is a good result: the
probability of obtaining the experimental curves by pure chance is negligible.
\figurename~\ref{fig:simfixp} shows two instances of successful simulations.
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_shortgenome_N_1e4_epitopes_example_longer}}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_N_1e4_3ingredients_3rddel_envnonenv_stopall}}
\caption{Simulations show that the suppression of fixation probability can be
generated by linkage to sweeping nonsynonymous alleles nearby. Two possible
scenarios are competition between escape mutants (left panel) and time-dependent
selection due to immune sytem recognition (right panel).}
\label{fig:simfixp}
\end{center}
\end{figure}

First of all, in order for $P_f$ to be reduced to levels seen in the data, most
synonymous mutants habe to be slightly deleterious. The reason is that if
synonymous alleles are an even mixture of deleterious and neutral ones, the
latter are much more likely to be carried to high frequencies by genetic draft,
because they present no burden to escape mutations. The synonymous $P_f$ is then
dominated by these neutral alleles and the depression induced by the few
deleterious ones becomes undetectable.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitch-hike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym}
\usepackage{color}
\usepackage{subfig}
\usepackage{graphicx}
%\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\maketitle

@@

HIV evolves rapidly within a single host during the course of the infection. This evolution is driven by strong selection imposed by the host immune system via killer T cells (CTLs) and neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high mutation rate of HIV \citep{mansky}.
When the host develops a CTL or AB response against a particular viral epitope, mutations that reduce or prevent recognition of the epitope frequently emerge. Escape mutations in epitopes targeted by CTLs typically evolve early in infection and spread rapidly through the population \citep{McMichael}. Later in infection, the most rapidly evolving part of the HIV genome are the so called variable loops of the envelope proteins that need to avoid recognition by neutralizing anti bodies. Mutation in \env~ spread through the population within a few months (see \figurename~\ref{fig:aft}, solid lines). During chronic infection, the (Malthusian) effect size of this beneficial
mutations is of the order of $s_e \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are selected for their effect on the amino acid sequence of the viral proteins. The viral genome, however, needs to meet additional constraints such as efficient processing and translation, nuclear export, and packaging into the viral capsid which operate on the RNA level. Purifying selection beyond the protein sequence is therefore expected, while it seems reasonable that the bulk of positive selection through the immune system should be restricted to amino acid sequences. A couple important RNA elements  are well characterized. For example, within \env{} a certain RNA sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. Recent studies
have shown that genetically engineered HIV strains with skewed codon usage bias
(CUB) patterns towards more or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.

INFLUENZA PSEUDO VACCINE.

SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE FOR SUPPLEMENT?

Despite evidence for functional importance of specific RNA sequences, synonymous mutations are commonly used as approximate neutral markers in studies of viral evolution. Neutral markers allow to make inference about the stochastic forces driving evolution \citep{smth}. Here, we characterize the dynamics of synonymous mutations in \env{} and show that a substantial fraction of these mutations are deleterious. The central quantity we investigate is the probability of fixation of a mutation, conditional on its population frequency. Even though the synonymous mutations are deleterious and cannot be used as neutral markers, we show the degree to which they hitch-hike with nearby non-synonymous mutations is very informative. Extending the analysis of fixation probabilities to the non-synonymous mutations, we show that time dependent selection or strong competition of escape mutations inside the same epitope are necessary to explain the observed patterns of fixation and loss.

One simple way to assess the neutrality of synonymous mutations is to look at
their level of conservation. Deleterious mutations at functional sites are
expected to be absent or rare across the viral population; vice versa, mutant
alleles that reach high frequencies are expected to be neutral. If genetic sites
are independent, the equilibrium frequency of a deleterious allele with fitness
$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
neutral alleles have no equilibrium frequency and can slowly fix via genetic
drift~\citep{ewens_mathematical_2004}.
t If the focal synonymous mutant is linked
to another nonneutral allele, however, its frequency is the result of the
combined fitness effects of both sites, and simple conservation-level analyses
fail. Since recombination in HIV is known to happen
rarely~\citep{neher_recombination_2010, batorsky_estimate_2011}, the genetic
context of the synonymous change at hand must be taken into account. Our
results underline the importance of the latter scenario for intrapatient HIV
evolution.

\section{Results}
A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to spread through the population and fix, while it is lost with probability $1-\nu$. This is a simple consequence of the fact that exactly one of present $N$ individuals will be the common ancestor of the entire population at a particular locus and this ancestor has a probability $\nu$ of carrying this mutations, see illustration in \FIG{illustration}. Deleterious or beneficial mutations, in contrast, should fix less or more often, respectively. Time series sequence data therefore suggest a simple way to investigate average properties of different classes of mutations.

\paragraph{Synonymous mutations in \env, C2-V5 are mostly deleterious}

\FIG{aft} shows time series data of the frequencies of all mutations observed \env, C2-V5, in patient ??\citep{shankarappa_consistent_1999,liu_selection_2006}. Despite many synonymous mutations reaching high frequency, very few fix. This observation in further quantified in panels ? and ?, that stratify the data of ?? patients (see methods) according to the frequency at which different mutations are observed. Considering all mutations in a frequency interval $\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency 1, frequency 0, or at intermediate frequency at later time points $t_f$. Plotting these fixed, lost, and polymorphic fraction against the time interval $t_f-t_i$, we see that most synonymous mutations segregate for roughly one year and are lost  much more frequently than expected. The ultimate probability of loss or fixation is shown as a function of the initial frequency $\nu_0$ in panel ??. In contrast to synonymous mutations, the non-synonymous seem to follow more a less the neutral expectation -- a point to which we will come back below.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
\caption{Synonymous mutations rarely fix in \env, C2-V5: Panel A shows mutation 
frequency trajectories observed in patient ?? \cite{shankarappa_consistent_1999};  Nonsynonymous
and synonymous mutations are shown as solid and dashed lines, respectively. 
Colors indicate the position of the site along the C2-V5 region (red to blue) ADD COLORBAR. MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN SEPARATELY.
While non-synonymous mutations frequently fix, very few synonymous mutations do even though they are frequently observed at intermediate frequencies. Panel B quantifies time course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous mutations that fix as a function of intermediate frequency $\nu_0$.
}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire \env~gene of ?? patients at ?? time points with ??-?? sequences each. Repeating the above analysis separately on the C2-V5 region studied above and the remainder of \env~ reveal strikingly different behavior inside and outside the hyper variable region. Within C2-V5, this data fully confirms the observations made in the data set by \citet{shankarappa_consistent_1999}. In the remainder of \env, however, observed synonymous mutations behave as if they were neutral; see \FIG{fixp}.

ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL? DOES LOSS/FIX CORRELATE WITH CONSERVATION. CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION.

These observations suggest that many of the synonymous mutations in the part of \env~that includes the hyper variable regions are deleterious, while outside this regions only roughly neutral mutations are polymorphic.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}}
\caption{Left panel: fixation probability of derived synonymous alleles is strongly
suppressed in C2-V5 versus other parts of the {\it env} gene, and of
nonsynonymous ones.
Right panel: especially hard is fixation of new alleles in conserved regions flanking the V
loops. The black dashed line is the prediction from neutral
theory, for comparison purposes. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

\paragraph{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
One possible {\it a priori} explanation for lack of fixation of synonymous mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA secondary structures are relevant for HIV replication,
mutations in nucleotides involved in those base pairs are expected  to be deleterious and to revert preferentially. 
Many functionally important secondary structure elements have been characterized, including  the  \rev{} response element (RRE) to enhance nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}. It has been suggested early on that parts of the viral genome that has the potential to form stems as better conserved that the remainder \citep{forsdyke_reciprocal_1995}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has been measured using the SHAPE assay (a biochemical reaction preferentially altering unpaired bases) \citep{watts_architecture_2009}. The SHAPE assay has shown that the variable regions V1 to V5 tend to be unpaired, while the conserved regions between those variable regions form stems. We partition all synonymous alleles observed
at intermediate frequencies above 10-15\% depending on their final destiny
(fixation or extinction). Subsequently, we align our sequences to the reference
NL4-3 strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
reactivities. As shown in \figurename~\ref{fig:SHAPE} (left panel) in a
cumulative histogram, the reactivity of fixed alleles are systematically larger
than of alleles that are doomed to extinction. In other words, alleles that are
likely to be breaking RNA helices are also more likely to revert and finally be
lost from the population. We then split the synonymous mutations in the C2-V5 region further into conserved and variable regions and found that the biggest depression in fixation probability is observed in the conserved stems, while the variable loops show little deviations from the neutral signature; see \FIG{SHAPE}B.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\paragraph{Deleterious mutations are brought to high frequency by hitch-hiking}
While the observation that some fraction of synonymous mutations is deleterious is not unexpected, it seems odd that we observe them at high population frequency -- at least in some regions of the genome. The region of \env~in which we observe deleterious mutations at high frequency is special in that it undergoes frequent adaptive changes to evade recognition by neutralizing antibodies \cite{Williamson}. Due to the limited amount of recombination in HIV \cite{neher_recombination_2010,batorsky_2011}, deleterious mutations that are linked to adaptive variants can reach high frequency \citep{maynard_smith}.

The potential for hitch-hiking is already apparent from the allele frequency trajectories in \FIG{aft}, where many mutations appear to change rapidly in frequency as a flock. Deleterious synonymous mutations can be amplified exponentially by selection on linked nonsynonymous sites, a process known as {\it genetic draft}~\citep{neher_genetic_2011}. In order to be advected to high frequency by a linked adaptive mutation, the deleterious effect of the mutation has to be substantially smaller than the adaptive effect. The latter was estimated to be on the order of $s_a = 0.01$ per day. The approximate magnitude of the deleterious effects can be estimated from \FIG{fixtimes}, that shows the distribution of times for synonymous alleles to reach the fix or get lost starting from intermediate frequencies. The typical time to loss is of the order of 500 days. If this loss is driven by the deleterious effect of the mutation, this corresponds to deleterious effects of roughly $0.002$ per day.

To get a better idea of the range of parameters that are compatible with the observations and our interpretation, we  perform computer simulations of evolving viral populations under selection and rare recombination. For this purpose, we use the recently published package FFPopSim, which includes a module dedicated to intra-patient
HIV evolution~\citep{zanini_ffpopsim:_2012}. We analyze many combinations of
parameters such as population size, recombination rate, selection coefficient
and density of escape mutations, deleterious effect of synonymous mutation.

The main result of the simulations is that genetic draft can indeed bring weakly deleterious mutations to high frequencies and result in a dependence of the fixation probability on initial frequency that is compatible with observations. We quantify the reduction in fixation probability by the area under the diagonal.... Since neutral mutations are much more likely to rise to high frequency than deleterious ones, the majority of the synonymous mutations needs to be slightly deleterious observe a significant reduction of $p_{fix}$. Furthermore, the two crucial parameters that control the fixation probability
are the following: (a) the deleterious effects of hitchhikers compared to
the beneficial effects of escape mutants, and (b) the density of escape
mutations. Intuitively, a higher density of escape mutations (i.e., epitopes)
enables a larger degree of genetic draft, because escape mutations start to
combine and their effects add up. In \figurename~\ref{fig:simheat} (left panel),
we show that this is indeed the case in simulations.

SHOW THAT THE DEPRESSION WORKS.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}}
\caption{Simulations on the escape competition scenario show that the density of
 selective sweeps and the size of the deleterious effects of synonymous
 mutations are the main driving forces of the phenomenon. A convex fixation
 probability is recovered, as seen in the data, along the diagonal (left panel):
 more dense sweeps can support more deleterious linked mutations. The density of
 sweeps is limited, however, by the nonsynonymous fixation probability, which is
 quite close to neutrality (right panel). Moreover, strong competition between
 escape mutants is required, so that several escape mutants are ``found'' by HIV
within a few months of antibody production.}
\label{fig:simheat}
\end{center}
\end{figure}
